% !TeX encoding = UTF-8
% !TeX program = xelatex
% ============================================================================
% NSL-KDD 网络入侵检测系统 —— 课程论文 (CJC 模板)
% 编译: latexmk -xelatex cjc-main.tex  或  make cjc
% ============================================================================

\documentclass{cjc}

\usepackage{booktabs}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{siunitx}
\usepackage{graphicx}
\usepackage{float}

\classsetup{
  title        = {基于NSL-KDD数据集的网络入侵检测研究——传统机器学习与深度学习方法对比},
  title*       = {Network Intrusion Detection Based on NSL-KDD Dataset: A Comparative Study of Traditional Machine Learning and Deep Learning Methods},
  authors      = {
    author1 = {
      name         = {刘卫},
      name*        = {LIU Wei},
      affiliations = {aff1},
      biography    = {},
      biography*   = {},
      email        = {},
    },
    author2 = {
      name         = {查恩鹏},
      name*        = {ZHA En-Peng},
      affiliations = {aff1},
      biography    = {},
      biography*   = {},
      email        = {},
    },
    author3 = {
      name         = {陈安旭},
      name*        = {CHEN An-Xu},
      affiliations = {aff1},
      biography    = {},
      biography*   = {},
      email        = {},
    },
  },
  affiliations = {
    aff1 = {
      name  = {青海大学\ 计算机技术与应用学院, 青海\ 西宁\ 810016},
      name* = {School of Computer Technology and Applications, Qinghai University, Xining 810016, China},
    },
  },
  abstract     = {
    本文基于NSL-KDD数据集（训练集125,973条，测试集22,544条），构建了一个完整的网络入侵检测系统（IDS）。
    通过特征工程将原始41维特征降维至20维关键特征，对比评估了10种模型（包括决策树、随机森林等传统方法，
    以及MLP、CNN、LSTM等深度学习方法）在二分类和多分类任务上的性能。
    实验结果表明，在验证集上神经网络方法显著优于传统机器学习方法，MLP的AUC达到0.999；
    但在统一测试集上两者性能差距缩小——MLP的F1值约为0.72，决策树约为0.76。
    多分类任务受测试集15种训练未见攻击和小样本类别的双重制约，全量准确率上限约10\%。
    本文系统分析了各模型的适用场景与局限性，为NSL-KDD入侵检测任务提供了全面的实验基准和方法论参考。
  },
  abstract*    = {
    This paper presents a comprehensive network intrusion detection system (IDS) based on the NSL-KDD dataset
    (125,973 training samples, 22,544 test samples). Through feature engineering, the original 41-dimensional
    feature space is reduced to 20 key features. A total of 10 models are systematically compared, including
    traditional machine learning methods (Decision Tree, Random Forest) and deep learning methods (MLP, CNN, LSTM),
    evaluated on both binary classification and multi-class attack recognition tasks. Experimental results show
    that deep learning methods significantly outperform traditional methods on the validation set, with MLP achieving
    an AUC of 0.999. However, on the unified test set, the performance gap narrows considerably—MLP achieves an F1
    score of approximately 0.72, while Decision Tree reaches 0.76. Multi-class classification is constrained by
    15 unseen attack types in the test set and extreme class imbalance, with full accuracy capped at approximately
    10\%. This paper provides a comprehensive experimental benchmark and methodological reference for NSL-KDD
    intrusion detection research.
  },
  keywords     = {网络入侵检测, NSL-KDD, 机器学习, 深度学习, 决策树, 随机森林, 对比分析},
  keywords*    = {network intrusion detection, NSL-KDD, machine learning, deep learning, decision tree, random forest, comparative analysis},
  clc          = {},
  doi          = {},
}

\newcommand\dif{\mathop{}\!\mathrm{d}}

% Graphics path
\graphicspath{{../outputs/figures/}}

% hyperref always loaded last in preamble
\usepackage{hyperref}

% Course assignment: suppress journal-specific header/footer on first page
\fancypagestyle{title}{%
  \fancyhf{}%
  \renewcommand{\headrulewidth}{0pt}%
}
% Also suppress running header on all pages (keep page number)
\fancypagestyle{plain}{%
  \fancyhf{}%
  \renewcommand{\headrulewidth}{0pt}%
  \fancyfoot[C]{\thepage}%
}
\pagestyle{plain}


\begin{document}

\maketitle

% ============================
% Chapters
% ============================
\input{chapters/ch1-intro}
\input{chapters/ch2-preprocess}
\input{chapters/ch3-ml}
\input{chapters/ch4-dl}
\input{chapters/ch5-comparison}
\input{chapters/ch6-conclusion}

% ============================
% Acknowledgments
% ============================
\begin{acknowledgments}
  本文工作得益于NSL-KDD数据集的开放共享以及scikit-learn、PyTorch等开源工具链的支持，在此表示感谢。
\end{acknowledgments}

% ============================
% References
% ============================
\bibliographystyle{cjc}
\bibliography{cjc-refs}


% ============================
% Background (English)
% ============================
\input{background}


% ============================
% Appendix
% ============================
\newpage
\appendix
\input{appendix}

\end{document}
